---
title: "0_SettingBoundaries"
author: "TP DuBose"
date: "12/8/2022"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse); library(sf); library(maps)
library(vegan); library(ggrepel); library(cowplot); library(MetBrewer); library(knitr)
pot_list<-read.csv('../AnuranMNMs/ATraiU 2.0/ATraiU2_summary_values_2022SEPT.csv') %>% pull(species)
```

## Species List

The goal is to get 10-20 species that can be compared across different climate change vulnerability assessments. We want to make sure that a good portion of each species range is within our study region, the Southeast United States, that they have unique traits when compared to congenerics / other anurans, and we capture a taxonomic gradient. 

So in a meeting on Dec 7, Meryl, Chloe and Traci identified nine species that would likely be great to run for the CCVAs based on their uniqueness and amount of range within the SE region. These species have high trait completeness for needed MNM traits (at least 3 of Tmerge, Tpref, CTmin and CTmax). These nine species are: `r wants<-c('Anaxyrus americanus', 'Anaxyrus fowleri','Scaphiopus holbrookii','Gastrophryne carolinensis','Lithobates clamitans','Lithobates catesbeianus','Dryophytes chrysoscelis','Dryophytes versicolor','Pseudacris crucifer')`

To expand this species list, I first approximate how much of each species area of occurrence is within the approximate SE region. 

```{r}
# pull in all the occurrence points used to calculate the RCS
anuran_pts<-NULL
for(s in pot_list[pot_list != 'Lithobates sevosus']){
  anuran_pts<-bind_rows(anuran_pts,
                        # for each species occurrences
                        geometry=read.csv(paste0('../National-RCS/data/occ_data/',s,'_20210330.csv')) %>% 
                        # only keep this information about the occurrence point
                          dplyr::select(source, species, Longitude, Latitude,
                                        coordinateUncertaintyInMeters, coordinatePrecision, elevation, 
                                        day, month, year) %>%
                          # convert it from a tibble to a sf
                          st_as_sf(coords=c('Longitude','Latitude'), 
                                   crs=st_crs("+proj=longlat +ellps=WGS84 +datum=WGS84")) %>%
                          # transform it from WGS84 to Alber's 
                          st_transform(st_crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=km +no_defs")) %>%
                          # take the union, which makes a multipoint objects
                          st_union() %>%
                          # and add a 1km buffer to the points
                          st_buffer(1) %>% 
                          # calculate the entire area, and rename the species and geometry column
                          bind_cols(geometry=., species=s, area=st_area(.))) 
}

# bind_cols removed the sf, so this is just reminding r that this is an sf in Albers crs
anuran_pts <- anuran_pts %>%
  st_as_sf(crs=st_crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=km +no_defs"))
```

For ease, we are using the spatial extent used to build the species list for the ATraiU-TaDS database. These values are thus approximate and should only be used for this decision process.

```{r crop aoo, warning=F, message=F}
# use the maps package to build build a shapefile
se<-st_as_sf(map('state',c('mississippi', 'georgia', 'florida', 'south carolina', 
                           'north carolina', 'virginia', 'alabama', 'tennessee'), 
                 fill=T, plot=F)) 
# make it one polygon and transform to Albers
sf::sf_use_s2(FALSE)
se.u<-se %>% st_make_valid() %>% st_union() %>% st_transform(st_crs(anuran_pts))
# crop species AOO to the region
se.aoos<-st_crop(anuran_pts, se.u) %>%
   bind_cols(se.area=st_area(.)) # measure the area within the region

# use the AOO area and the SE area to calculate 
# the percent of the species AOO within our region
poss.spp<-se.aoos %>%
  as_tibble() %>%
  mutate(per.inreg=round(as.numeric(se.area)/as.numeric(area)*100,1))
```

Just to note, the species already identified above range from `r poss.spp %>% filter(species %in% wants) %>% pull(per.inreg) %>% max()`% to `r poss.spp %>% filter(species %in% wants) %>% pull(per.inreg) %>% min()`% of their range within the general southeastern United States.

Then I pull in the trait data to quantify how many were found within the literature for the 37 possible species we could include. I then join this with the inputed data so that I have the number of and which traits were found within the literature, with the traits we can use in the MNM models. Because *Lithobates kauffeldi* was not included in the Jetz & Pyron phylogeny used to inform trait imputation, we cannot include it within our species list. The literature-found and inputed traits will be used to conduct a PCA to look at trait uniqueness among taxa. 

```{r trait availability}
trait_avail<-read.csv('../AnuranMNMs/ATraiU 2.0/ATraiU2_summary_values_2022SEPT.csv') %>%
  select(species, Mass:Tmerge) %>% select(-Tforage_optim) %>%
  group_by(species) %>%
  pivot_longer(-species) %>%
  filter(!is.na(value)) %>% 
  dplyr::summarize(n_traits=n(),
            traits=paste(name, collapse=', '), .groups='drop') %>%
  left_join(read.csv('../AnuranMNMs/input_data/MNM_traits_SEspp.csv'),
            by='species')
```

I first explored the variation in traits and identified outliers that were not included in the first list outlined in the 12/7 meeting. 

```{r trait biplots, message=F, warning=F, fig.width=6.5, fig.height=2.7}
maybes<-c('Dryophytes gratiosus', 'Lithobates areolatus', 'Dryophytes cinereus', 'Pseudacris fouquettei',
          'Lithobates sphenocephalus', 'Acris gryllus', 'Lithobates grylio')
trait_avail %>%
  left_join(poss.spp %>% select(species, per.inreg), by='species') %>%
  select(-traits, -s.fossorial, -s.arboreal, -nocturnal, -diurnal, -crepuscular) %>%
  pivot_longer(-c(n_traits, species, per.inreg)) %>%
  filter(!(name %in% c('Tforage_max','Tforage_min', 'Tbask'))) %>%
  mutate(category=factor(case_when(species %in% wants ~ 'Included',
                            species %in% maybes ~ 'Likely',
                            per.inreg < 30 ~ 'Range Edge', 
                            T ~ 'Possible'), levels=c('Included','Likely','Possible','Range Edge'))) %>%
  ggplot()+
  geom_point(aes(x=value, y=per.inreg, color=category), size=2, alpha=.5) +
  facet_wrap(~name, scales='free_x')+
  theme_classic()+
  scale_color_viridis_d()+
  scale_y_continuous('Percent AOO within SE region')+
  theme(legend.position=c(.87,.25),
        strip.background = element_blank())
```

I then conducted a quick PCA to investigate I was capturing unique trait combinations. There are likely more elegant ways to do this (functional uniqueness) but I thought this sufficient for our needs. The names of the species that we will include (per 12/7 meeting) and those with less that approximately 30% of their range within the SE region are not plotted below to aid in visualization. 

```{r trait pca var, fig.width=3, fig.height=2}
trait.pca<-rda(trait_avail[-24, c(5:7,11)])
# investigate variation on different axis
screeplot(trait.pca); abline(h = 1, col="red", lty=5)
pca.sum<-summary(trait.pca)
```


```{r pca plot, echo=F, fig.width=6.5, fig.height=4}
pca.spp.df<-pca.sum$sites %>% as_tibble() %>% bind_cols(species=trait_avail[-24,]$species) %>%
  left_join(poss.spp, by='species') %>%
  mutate(category=factor(case_when(species %in% wants ~ 'Included',
                            species %in% maybes ~ 'Likely',
                            per.inreg < 30 ~ 'Range Edge', 
                            T ~ 'Possible'), levels=c('Included','Likely','Possible','Range Edge')),
         pltsp=paste0(substr(species, 1,1), '. ', sub(".*? ", "", species)))
pca.env.df<-pca.sum$species %>% as_tibble() %>% mutate(variable=rownames(pca.sum$species))
pca.plt<-ggplot()+
  geom_text_repel(data=pca.spp.df %>% filter(category == 'Likely' | category == 'Possible'),
             aes(x=PC1, y=PC2, label=pltsp), size=3,
             max.overlaps = 10, min.segment.length = 0)+
  geom_point(data=pca.spp.df,
             aes(x=PC1, y=PC2, color=category), size=3, alpha=.7)+
    geom_segment(data=pca.env.df, 
            aes(x=0, y=0, xend=PC1, yend=PC2))+
  geom_text_repel(dat=pca.env.df, 
            aes(x=PC1, y=PC2, label=variable), color='blue')+
  scale_color_viridis_d()+
  theme_classic()+
  theme(legend.position='bottom')
pca.legend<-get_legend(pca.plt)
plot_grid(
  plot_grid(pca.plt + theme(legend.position='none'),
            ggplot()+
              geom_text_repel(data=pca.spp.df %>% filter(category == 'Likely' | category == 'Possible'),
                              aes(x=PC3, y=PC2, label=pltsp), size=3,
                              max.overlaps = 10, min.segment.length = 0)+
              geom_point(data=pca.spp.df,
                         aes(x=PC3, y=PC2, color=category), size=3, alpha=.7)+
              geom_segment(data=pca.env.df, aes(x=0, y=0, xend=PC3, yend=PC2))+
              geom_text_repel(dat=pca.env.df, 
                              aes(x=PC3, y=PC2, label=variable), color='blue')+
              scale_color_viridis_d()+
              theme_classic()+
              theme(legend.position='none')),
  pca.legend, ncol=1, rel_heights = c(.9,.07))
```

As a way to quantify this, I calculate the rank score for each species on each axis and the rank total AOO. I then subtract 18 from the rank (as the middle rank) and take the absolute value, so that species with higher rank values represent the ends of the spectrum of trait space or range size. 

```{r}
spp.rnks<-pca.spp.df %>%
  select(species, starts_with('PC'), per.inreg, area) %>%
  select(-PC4) %>%
  mutate(across(-c(species, per.inreg), ~abs(rank(.x)-18)),
         category=factor(case_when(species %in% wants ~ 'Included',
                                   species %in% maybes ~ 'Likely',
                                   per.inreg < 30 ~ 'Range Edge', 
                                   T ~ 'Possible'),
                         levels=c('Included','Likely','Possible','Range Edge'))) %>%
  rowwise() %>% mutate(mean.rank=mean(c(PC1, PC2, PC3, area))) %>% ungroup() %>%
  arrange(category) 
```

```{r rank tables, echo=F}
kable(spp.rnks %>%
        select(category, species, starts_with('PC'), area, per.inreg, mean.rank),
      caption='Rank uniqueness in trait space and range size. Higher values means the species is at the end of the spectrum (e.g., very small AOO or very large AOO). Species in the category Included were discussed on 12/7, Likely were identified by TPD in this document, Possible could be added if needed but might be similar to already included species, and Range Edge has less than 30% of its AOO within the SE region. PC1 is Tmerge, PC2 splits CTmax and CTmin, and PC3 splits CTmax and Tpref.',
      col.names = c('Category','Species', 'PC1','PC2','PC3', 
                    'AOO', '% in SE', 'Mean Rank'))
kable(spp.rnks %>% group_by(category) %>% 
  summarize(mean.rnk=mean(mean.rank),
            min.rnk=min(mean.rank), 
            max.rnk=max(mean.rank), .groups='drop'),
  caption='Summarized Rank Values for the four categories outlined. Note that Included and Likely is typically higher than the two other categories. The minimum rank in likely is Acris gryllus and adds a genera to our taxa list.',
  col.names = c('Suggested Category','Mean Rank', 'Minimum Rank', 'Maximum Rank'))
#write.csv('spp.rnks''.n , 'possible_species_list.csv', row.names = F)
```

Below, I present where the species in the proposed list (Included + Likely, points) align among the trait distributions of the 36 possible species (curves). 

```{r trait gradient, echo=F, warning=F, message=F}
trait_avail %>%
  select(species, Mass, CTmax, CTmin, Tpref, Tmerge) %>%
  pivot_longer(-species) %>%
  mutate(category=factor(case_when(species %in% wants ~ 'Included',
                            species %in% maybes ~ 'Likely'),
                            levels=c('Included','Likely'))) %>%
  #filter(!is.na(category)) %>%
  ggplot(aes(x=name, y=value))+
  geom_violin( )+
  geom_point(data=. %>% filter(species %in% c(wants, maybes)),
             aes(color=category),
             size=2.5, alpha=0.5)+
  facet_wrap(~name, scales='free', nrow=1)+
  scale_color_viridis_d(option='C', end=.2, begin=.8)+
  theme_classic()+
  labs(x='Variable', y='Gradient (values)')+
  theme(strip.background = element_blank(),
        axis.title.x = element_blank(),
        axis.text.x = element_blank(),
        axis.line.x = element_blank(),
        axis.ticks.x = element_blank())
```

# Spatial Extent of the Comparison

Four possible ways to define the extent of the southeastern United States are: state geopolitical boundaries, USGS agency boundaries, ecoregion boundaries, and watershed boundaries. Below are maps to help us explore which spatial extent works best for our climate change vulnerability assessment. 

In building the ATraiU-TaDS database, we used the following state geopolitical boundaries as the extent of the southeastern United States. Another option is The Department of Interior identified 9 [Unified Regional Boundaries](https://www.doi.gov/employees/reorg/unified-regional-boundaries) based on watersheds that could be used across DOI agencies, of which there is a 'South Atlantic-Gulf' region. Mississippi is included in the Mississippi Basin (Louisiana to Iowa included) and Virigina is a part of the North Atlantic-Appalachian Region.

```{r se map, echo=F, fig.height=3, fig.width=6.5}
plot_grid(ggplot()+geom_sf(data=se)+ theme_void()+ggtitle('ATraiU-TaDS'),
          ggplot()+geom_sf(data=se %>% filter(ID != 'virginia', ID !='mississippi'))+
            theme_void()+ggtitle('DOI Unified Region'))
```

More ecological boundaries include EPA ecoregions or USGS watershed boundaries. Below the [EPA Level III ecoregions](https://www.epa.gov/eco-research/level-iii-and-iv-ecoregions-continental-united-states) are presented. We could use Level IV if we need smaller regions. The [USGS HUC 2 watershed boundary](https://www.usgs.gov/media/images/watershed-boundary-dataset-subregions-map) for the '03' region might be a good ecological option that captures what we typically consider the southeastern United States. 

```{r eco data}
huc4s<-read_sf('C:/Users/Owner/Documents/GISfile/WBD_National_GDB/WBD_National_GDB.gdb',
        layer='WBDHU4') %>%
  mutate(huc2=substr(huc4, 1,2)) %>%
  filter(huc2 %in% c('08','06','03'))
epa_eco<-read_sf(dsn = '../us_eco_l3') %>%
  filter(grepl('8.3', L2_KEY)|grepl('8.4', L2_KEY)|grepl('8.5', L2_KEY)|US_L3CODE == '76',
         # these level three regions are removed but are within the 'Eastern Forests' ecoregion'
         !(US_L3NAME %in% c('Atlantic Coastal Pine Barrens', 
                            'East Central Texas Plains', 'South Central Plains',
                            'Ozark Highlands','Ouachita Mountains','Boston Mountains','Arkansas Valley',
                            'Western Allegheny Plateau'))) 

```

```{r eco maps, fig.width=6.5, fig.height=3, echo=F}
library(MetBrewer)
plot_grid(epa_eco %>%
  ggplot()+geom_sf(aes(fill=US_L3NAME))+geom_sf(data=se, fill=NA, color='white')+
  theme_void()+
    theme(legend.text = element_text(size=10),
        legend.title=element_text(size=10),
        legend.key.size = unit(.4, 'cm'))+
    scale_fill_manual('Ecoregion', values=c(met.brewer('Signac')[-14], 'light grey'))+
    ggtitle('EPA Ecoregions Level III'),
ggplot()+
  geom_sf(data=huc4s, aes(fill=huc2))+
  geom_sf(data=se, fill=NA, color='white')+
  theme_void()+scale_fill_grey('HUC 2')+
  ggtitle('USGS HUC 4')+
  theme(legend.position=c(.8,.4)),
rel_widths = c(.7,.4)) 
```

