---
title: "regional RCS calculation"
author: "TP DuBose"
date: "1/5/2023"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Hmisc)
library(sf); library(tidyverse); library(scales) # load necessary libraries.
library(terra); library(exactextractr)
library(cowplot) #for plotting maps later
```

```{r data prep}
# Set data paths -----
# data used
PATH_iucn_ranges <-'/home/tracidubose/RCS_Anuran_input/ANURA/' #IUCN range maps downloaded 10/2020
PATH_HUC12 <- '/home/tracidubose/huc12_wgs84_sf_02010323.rds'
PATH_FocalSpecies_OccurrenceData <- "/home/tracidubose/RCS_Anuran_input/Anuran Occurrence Data/"

# Set spatial coordinate reference system (CRS).
crs.geo <- st_crs("+proj=longlat +ellps=WGS84 +datum=WGS84")
crs.albers <- st_crs("+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=37.5 +lon_0=-96 +x_0=0 +y_0=0 +ellps=GRS80 +datum=NAD83 +units=km +no_defs")

# Read in fixed data -----
# Focal species & Spatial Extent determined in 0_SettingTaxSpaceExtents.Rmd
anuran.taxa<-c("Acris gryllus", "Acris crepitans",
               "Anaxyrus terrestris", "Anaxyrus fowleri", "Anaxyrus quercicus", 
               "Dryophytes avivoca",  "Dryophytes andersonii", "Dryophytes cinereus", 
               "Dryophytes chrysoscelis", "Dryophytes femoralis", "Dryophytes gratiosus",
               "Dryophytes squirellus",
               "Gastrophryne carolinensis",
               "Lithobates capito", "Lithobates grylio", "Lithobates okaloosae", 
               "Lithobates heckscheri", "Lithobates sphenocephalus", "Lithobates areolatus", 
               "Lithobates virgatipes",
               "Pseudacris ornata", "Pseudacris feriarum", "Pseudacris brachyphona",
               "Pseudacris fouquettei",
               "Scaphiopus holbrookii")
#IUCN maps for native ranges
iucn_anuran<-read_sf(PATH_iucn_ranges) %>%
  filter(binomial %in% anuran.taxa) %>% 
  st_transform(crs.albers) 
# prism file location
PATH_prism<-'/home/tracidubose/prism_files/'
PATH_climout<- '/home/tracidubose/b_RegionalRCS/climate_outliers/'

calcmode <- function(a){  
  vector <- unique(a)  
  vector[which.max(tabulate(match(a, vector)))]} 

# climate outliers QA/QC
climate_outliers<-read.csv('RCS_rerun1207/climate_outlier_pts.csv')

# spatial extent
huc12s<-readRDS(PATH_HUC12) %>%
  mutate(huc2=substr(huc12, 1,2)) %>% 
  filter(huc2 %in% c("03","06","08"))
spat.ext<-readRDS('b_RegionalRCS/spatial_extent.rds') %>% st_transform(crs.albers) %>%
  st_buffer(.1)
ggplot() + geom_sf(data=spat.ext) + theme_void()
```

# Prep Climate Data

```{r}
library(prism)
clim.variables<-c('ppt','Tmax','Tmin')
spat.big<-spat.ext %>% st_buffer(100)
for(cm in clim.variables){
  options(prism.path = paste0("//home/tracidubose/prism_files/", cm, "/"))
  PRISM_data <- pd_stack(prism_archive_ls()) 
  spat.big<- st_transform(spat.big, crs(PRISM_data))
  PRISM_se<-crop(PRISM_data, st_bbox(spat.big))
  ter.prism<-rast(PRISM_se)
  names(ter.prism)<- gsub(paste0('PRISM_', str_to_lower(cm),'_stable_4kmM2_'), '',
                          gsub(paste0('PRISM_', str_to_lower(cm),'_stable_4kmM3_'), '',
                             gsub('_bil', '', names(ter.prism))))
  assign(paste0(cm,'.prism'), ter.prism)
}
rm(PRISM_data, spat.big, PRISM_se, ter.prism)
```

```{r}
head(Tmax.prism)[1:5]
huc.Tmax<-exact_extract(Tmax.prism, huc12s, 'mean') %>%
  bind_cols(huc12=as.character(huc12s$huc12))
head(huc.Tmax)[c(1:3,126)]
head(Tmin.prism)[1:5]
huc.Tmin<-exact_extract(Tmin.prism, huc12s, 'mean')%>%
  bind_cols(huc12=as.character(huc12s$huc12))
head(huc.Tmin)[c(1:3,126)]
head(ppt.prism)[1:5]
huc.ppt<-exact_extract(ppt.prism, huc12s, 'mean')%>%
  bind_cols(huc12=as.character(huc12s$huc12))
head(huc.ppt)[c(1:3,126)]
```


Once all the data is loaded, we first create a function that calculates area of occurrence at a specific spatial extent and specific grain size. The function below uses the usa_l48 shapefile as an anti.spatial.file to calculate point occurrences outside of the conterminous United States. It allows you to indicate whether you want to use watersheds or a specific buffering radius as your spatial grain size. You can also indicate what you want your output file to be named. 

```{r}
myfun<-function(x,index){bind_cols( rowname=as.character(index), x)}
find_area<-function(taxa, # which species to find the aoo for
                    spatial.file1, # which sp object to use for filtering/grain sizes
                    native_range=T, #do you want to exclude points >100km outside the native range
                    watershed=F, #are you using watersheds as grain sizes?
                    rad=1, #if buffering points, what should the buffer radius be?
                    clim.vars, #list of climate variables to check for outliers
                    watershed.name.column){
  if(!is.na(spatial.file1)){
    spatial.file<-get(spatial.file1)
    crs.set<-st_crs(spatial.file)}else{
      crs.set<-crs.albers
    }

  #read in sp occurrence data
  geodata<-read.csv(list.files(PATH_FocalSpecies_OccurrenceData, pattern=taxa, full.names = T)) %>%
    dplyr::select(family, genus, species, Longitude, Latitude, year, source, final.taxa) %>%
    filter(!is.na(year), year > 1894) %>% 
    st_as_sf(coords=c("Longitude","Latitude"), crs = crs.geo) %>% #convert to a sf dataframe
    rowid_to_column() # add this to align climate outliers table with this
  orig.nrow<-nrow(geodata)
  
  # restrict points to those found within IUCN native range
   if(native_range==T){
    nat_range <- iucn_anuran %>% filter(binomial %in% taxa) %>%
      st_union() %>% st_buffer(100) 
    geodata <- geodata %>% st_transform(crs.albers) %>% st_filter(nat_range)
  } 
  postNR.nrow<-nrow(geodata)
  
  #identify any climate outliers to exclude
   # create 1 km buffer
  #spp_BUF <- st_buffer(geodata, dist = 1) %>%
   # st_transform("EPSG:4269")
    # extract the values of interest from the raster for each polygon
  #clim_out<-NULL
  #for(env.r in clim.vars){
   # env.raster<-rast(paste0(PATH_prism,env.r,'_mean_usa.tif'))
    #clim_out_v<-spp_BUF %>% as_tibble() %>%
     # dplyr::select(-geometry) %>%
      #bind_cols(clim_value=exact_extract(env.raster, spp_BUF, fun='mean', progress=F)) %>%
      #mutate(mean_v=mean(clim_value),
       #    sd_v=sd(clim_value),
        #   zscore=(mean_v-clim_value)/sd_v) %>%
    #filter(abs(zscore) > 4)
  #clim_out <- bind_rows(clim_out, clim_out_v)
  #}
  #if(nrow(clim_out)!=0){
   # print(paste(length(unique(clim_out$rowid)), 'of', nrow(geodata), 'points removed as climate outliers for', env.r, 'and', taxa))
  # keep record of all excluded points
  #write.csv(clim_out, paste0(PATH_climout, taxa, '_climout.csv'))}
  #exclude climate outliers
  clim_out <- climate_outliers %>% filter(species == taxa)
  dat_sp<- geodata %>% 
    filter(!(rowid %in% clim_out$rowid))
  postCL.nrow <-nrow(dat_sp)
 
  # Calculate area occupied per species. 
  if(watershed==T){
    dat_sp <- st_transform(dat_sp, st_crs(spatial.file)) # ensuring transformations fit between spatial file and points
    dat_sp <- st_join(dat_sp, spatial.file) %>%
      filter(!is.na(huc12)) %>%
      arrange(desc(year)) %>%
      distinct(huc12, year, .keep_all=T)
     all.poly.nrow<-nrow(dat_sp) #identify howmany watersheds are occupied
    area.type<- paste0(spatial.file1,'_WS')
    used.nrow<-nrow(dat_sp)
    min.year<-min(dat_sp$year)
    area<-dat_sp %>% distinct(huc12) %>% pull(WSAREA) %>% sum() %>% as.numeric()
    
    # Climate Sensitivity based on the AOO calculated above
    CS_out<-NULL
    for(env.r in paste0('huc.',clim.vars)){
    env.raster<-get(env.r)
    CS_out_v <- dat_sp %>% as_tibble() %>% 
      select(-geometry) %>%
      left_join(env.raster, by="huc12") %>% 
      dplyr::select(huc12, year, states, starts_with('mean.')) %>%
      pivot_longer(starts_with('mean.')) %>%
        mutate(climate_year=gsub('mean.', '', name)) %>%
      filter(year >= climate_year, year-30 < climate_year) %>%
      group_by(huc12, year) %>%
      dplyr::summarize(mean_v=mean(value),
                       sd_v=sd(value), .groups='drop') %>%
      ungroup() 
      CS_out<-bind_rows(CS_out, 
                        CS_out_v %>% 
                          dplyr::summarize(modyear=calcmode(year),
                                           oldyear=sum(year < (2020-50)),
                                           clim=gsub('huc.','',env.r),
                                           Cmean=mean(mean_v),
                                           Csd=sd(mean_v))) %>%
        mutate(spp=taxa)
      if(spatial.file1=='huc12s'){
      write.csv(CS_out_v, paste0(taxa,'_', env.r, '_huc_climate_info.csv'))}
    }
  }else{
    # Generate 1km point buffers. You must use a projected CRS. For CRS, we use: USA Contiguous albers equal area. 
     area.type<- paste0(spatial.file1,'_',rad,'km')
     
     dat_sp <- st_transform(dat_sp, st_crs(spatial.file)) %>% #overwrites to the projection 
      st_filter(spatial.file) %>% # only keep points pts in the spatial extent
       arrange(desc(year)) %>% # keep newer years first when discarding overlaps
       st_transform(crs.albers) %>% st_buffer(1) # buffer by the requested radius
     #print(head(dat_sp))
  # calculate total area 
     area<-st_union(dat_sp) %>% st_area(.) %>% as.numeric()
     all.poly.nrow<-nrow(dat_sp)
     min.year<-min(dat_sp$year)
     # prep the data frame for climate variable extraction  
     dat_sp <- dat_sp %>%
       count(year) %>%
       rownames_to_column('rowname') %>%
       bind_cols(area=st_area(.)) %>%
       st_transform(st_crs(ppt.prism)) 
#     print(head(dat_sp))
     used.nrow<-nrow(dat_sp) # keep track of polygons used for climate after removal of small polys
     
    CS_out<-NULL
    for(env.r in paste0(clim.vars, '.prism')){
    env.raster<-get(env.r)

    env_vals_raw<-exact_extract(env.raster, dat_sp, progress=F, force_df=T)
    env_vals<-do.call('rbind',
                      mapply(myfun, env_vals_raw, seq_along(env_vals_raw), SIMPLIFY = F)) %>% 
      left_join(dat_sp %>% as_tibble() %>% select(-geometry), by='rowname') %>%
      dplyr::select(rowname, year, area, coverage_fraction, n, everything()) %>%
      pivot_longer(-c(rowname, year, area, coverage_fraction, n))
    if(nrow(filter(env_vals, is.na(value))) != 0){
      print(paste(nrow(filter(env_vals, is.na(value))), 'suspicious points for', env.r, 'and', taxa))
      print(head(filter(env_vals, is.na(value))))}
    CS_out_v <- env_vals %>%
      mutate(climate_year=gsub('mean.', '', name)) %>%
      filter(year >= climate_year, year-30 < climate_year) %>%
      group_by(rowname, year, area, n, coverage_fraction) %>%
      dplyr::summarize(nv=mean(value), .groups='drop') %>%
      distinct(rowname, coverage_fraction, nv, .keep_all=T) %>%
      ungroup() %>%
      dplyr::summarize(spp=taxa,
                   modyear=calcmode(year),
                   oldyear=sum(year < (2020-50)),
                   #clim=gsub('.prism','',env.r),
                   Cmean=wtd.mean(nv, weights=coverage_fraction, normwt = T),
                   Csd=sqrt(wtd.var(nv, weights=coverage_fraction, normwt = T)))
    CS_out<-bind_rows(CS_out, CS_out_v)
  }
  }  
  RCS_comp <- data.frame(scientific_name = taxa, 
                     area.type = area.type,
                     area.sqkm = area,
                     orig.nrow=orig.nrow,
                     postNR.nrow=postNR.nrow,
                     postCL.nrow=postCL.nrow,
                     postDif.nrow=all.poly.nrow,
                     used.nrow=used.nrow,
                     min.year=min.year,
                     stringsAsFactors = F) %>%
    bind_cols(CS_out %>%
                select(-spp) %>%
                        pivot_wider(names_from = clim,
                                    values_from=c("Cmean","Csd")))
  
  return(RCS_comp)
}
```

```{r}
# notes: need to investigate Anaxyrus quercicus
find_area(taxa="Acris crepitans",
                  spatial.file1="spat.ext", rad=5,
                  clim.vars = c('ppt','Tmax','Tmin'))

```



```{r}
library(maps)
se<-st_as_sf(map('state',c('mississippi', 'georgia', 'florida', 'south carolina', 
                           'north carolina', 'alabama'), 
                 fill=T, plot=F)) %>%
  st_transform(crs.albers)
al<-se %>% filter(ID == 'alabama')
fl<-se %>% filter(ID == 'florida')
sc<-se %>% filter(ID == 'south carolina')
nc<-se %>% filter(ID == 'north carolina')
ga<-se %>% filter(ID == 'georgia')
ms<-se %>% filter(ID == 'mississippi')
huc3<-huc12s %>% filter(huc2=="03")
huc6<-huc12s %>% filter(huc2=="06")
huc8<-huc12s %>% filter(huc2=="08")
```

We then apply this function at the differen grain sizes to create a single dataframe with out RCS components. 

```{r}
RCS_comp_all<-NULL
start.time<-Sys.time()
for(extent.type in c("spat.ext", "huc3","al")){
for(spp in anuran.taxa[c(1:2,5:7)]){
  buff1<-find_area(taxa=spp,
                  spatial.file1=extent.type, rad=1,
                  clim.vars = c('ppt','Tmax','Tmin'))
  buff5<-find_area(taxa=spp,
                  spatial.file1=extent.type, rad=5,
                  clim.vars = c('ppt','Tmax','Tmin'))
  huc3out<-find_area(spp, 
          "huc3", native_range=T, watershed=T, 
          clim.vars=c('ppt','Tmax','Tmin'))
  huc12all<-find_area("Acris gryllus", 
          "huc12s", native_range=T, watershed=T, 
          clim.vars=c('ppt','Tmax','Tmin'))

  RCS_comp_all<-bind_rows(RCS_comp_all, 
                          bind_rows(buff1, buff5, huc12all))
}
}
end.time<-Sys.time()
end.time-start.time
```

```{r}
RCS_comp_all %>% arrange(scientific_name)

```

```{r}
# non-moving window result
read.csv('RCS_data_mean_climate.csv') %>% filter(scientific_name%in% unique(RCS_comp_all$scientific_name))

```

```{r}

rcs_index<-function(x)(1-(x-min(x))/(max(x)-min(x)))
rcs_res<-RCS_comp_all %>%
  group_by(area.type) %>%
  select(scientific_name, area.type, area.sqkm, starts_with('Csd'))%>%
  mutate(across(where(is.numeric), ~rcs_index(.x))) %>%
  rowwise() %>%
  mutate(RCS=(area.sqkm+mean(c(Csd_ppt, Csd_Tmax, Csd_Tmin)))/2) %>%
  arrange(desc(RCS))
#RCS_comp_all %>% 
 #  select(scientific_name, area.type, area.sqkm, starts_with('Csd')) %>%
 #  arrange(desc(Csd_ppt))
```

```{r}
rcs_ord<-rcs_res %>% filter(grepl('spat.ext_5km', area.type)) %>% arrange(RCS) %>% pull(scientific_name)
rcs_res %>%
  mutate(spF=factor(scientific_name, levels=rcs_ord),
         ext.type=case_when(grepl('spat.ext', area.type)|grepl('huc12', area.type)~'entire extent',
                            substr(area.type,1,2) %in% c('al','nc','sc','ga','ms','fl')~ 'state extent',
                            substr(area.type, 1,3) =='huc'~'HUC2'),
          `Grain Size`=substr(area.type, nchar(area.type)-2, nchar(area.type))) %>%
  ggplot(aes(y=spF, x=RCS), alpha=0.5)+
  geom_point(aes(shape=`Grain Size`), size=2)+
  scale_x_reverse('Rarity and Climate Sensitivity Index') +
  theme_classic()+
  scale_shape_manual(values=c(1,2))+
  facet_wrap(~ext.type) +
  theme(axis.text.y = element_text(face='italic'),
        axis.title.y=element_blank(),
        panel.grid.major.y = element_line(color="lightgrey"))
```


